Computing the bicharacter of huge subspaces of polynomials

### 2019-04-30 article/article.tex
Add definitions in context section

%Notes
\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\newcommand{\pauline}[1]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue]{#1}}
\newcommand{\nicolas}[1]{\todo[linecolor=green,backgroundcolor=green!25,bordercolor=green]{#1}}

@@

	\section{Introduction}
	This package was developped at the occasion of a collaboration with François Bergeron to study the character of the space $H_{k,n}$ of diagonal harmonic polynomials on $k$ sets of $n$ variables, as a graded bi $GL_k$-$S_n$  module. The main achievement is a calculation of this character for $n=6$ and $r=5$ (and therefore any $r$) in 45 minutes (and ~15Gb), which led to conjectures on its general value.

	This is a potentially huge calculation ($H_{5,6}$ is of dimension $5.8 \cdot10^5$), involving polynomials of degree up to $15$ in $30$ variables with thousands of terms. So we had to combine several strategies that each reduced the time and/or memory complexity by one or two orders of magnitude. As much as possible, the code was written to be modular and reusable.

	\section{Contexte}
	Note: throughout this notebook, we implicitly use the Frobenius map, and identify $S_n$ and $GL_k$ characters with their associated symmetric functions.

	Consider the polynomial ring $P=\QQ[X]$ with $k$ sets of $n$ variables. To fix conventions, we think of $X$ as a matrix of variables:
	$$(X_{i,j})_{0\leq i<k, 0\leq j<n}$$

	The polynomial ring $P$ is endowed with an action of the symmetric group $S_n$ by diagonal permutation of the columns, and $GL_k$ by diagonal action on the columns. The invariant ring $P^{S_n}$ is also known as the ring of multisymmetric functions. The coinvariant ring $P_{S_n}$ is the quotient of $R$ by the ideal $\langle {P^{S_n}}^+ \rangle$. It inherits from $P$ a structure of graded $GL_k$-$S_n$ bimodule.

	The grading can be refined into a multigrading by rows. The multigrading is preserved by the action of $S_n$. The multigrading is not preserved by the action of $GL_k$, but instead carries exactly the information about the $GL_r$ character. It's thus equivalent to consider the $GL_r$-$S_n$ bicharacter or the $S_n$ multigraded character of $P_{S_n}$.

	When $k=1$, the action of $GL_1$ is trivial; $P_{S_n}$ is of dimension $n!$ and carries the usual graded representation of $S_n$. Its character is given by a Hall-Littlewood polynomial.
	When $k=2$, $P_{S_n}$ is of dimension $(n+1)^{n-1}$ ($=16807$ for $n=6$); as a consequence of Haiman's proof of the $n!$-conjecture, the bigraded character is a Macdonald polynomial.

@@

	We first need to compute in a quotient by an ideal.
	A standard approach would be to compute a Gröbner basis of this ideal.
	We haven't actually tried it, but given the number of variables and dimension of the quotient, we expect this calculation to be intractable.
	Beside

@@

	Once a basis of $\harmonics_{k,n}$ is constructed, there still remains to recover its bicharacter. For $GL_k$ that's given by the multigrading. For $S_n$ some additional work has to be done (e.g. computing the traces of the action of permutations of all cycle types).

@@

	Advantages:
	- Writing $N=:\sum N_D$, The complexity becomes $(\sum N_D^2) M$ instead of $(\sum N_D)^2M_D$ (in fact better.
	- The multidegrees, and therefore $GL_k$ character, can be directly read off the result.

@@

	The polarization operators commute with the action of $S_n$. Therefore,
	we can compute independently each $S_n$ isotypic component of $\harmonics_{k,n}$ by starting from a basis of harmonic polynomials that itself is adapted to its decomposition into isotypic components:
	$$\harmonics_{n}=:\bigoplus_\lambda \harmonics^\lambda\,.$$

@@

	Remark: For a fixed tableau $P$, denote by $\harmonics_k^P$ the space spanned by the higher Specht polynomials $(G_{P,Q})_Q$ under polarization. It's exactly the image of $\harmonics$ under the Young idempotent associated to $P$. For a given shape $\lambda$, the $\harmonics_k^P$ are all conjugate under $S_n$, and their direct sum is the isotypic component $\harmonics_k^\lambda$.

@@

>>>>

%Notes
\usepackage[colorinlistoftodos,prependcaption,textsize=small,textwidth=2.5cm]{todonotes}
\newcommand{\pauline}[1]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue]{#1}}
\newcommand{\nicolas}[1]{\todo[linecolor=green,backgroundcolor=green!25,bordercolor=green]{#1}}

@@

	\section{Introduction}
	This package was developped at the occasion of a collaboration with François Bergeron to study the character of the space $\harmonics_{k,n}$ of diagonal harmonic polynomials on $k$ sets of $n$ variables, as a graded bi $\GLr$-$\Sn$  module. The main achievement is a calculation of this character for $n=6$ and $r=5$ (and therefore any $r$) in 45 minutes (and ~15Gb), which led to conjectures on its general value.

	This is a potentially huge calculation ($\harmonics_{5,6}$ is of dimension $5.8 \cdot10^5$), involving polynomials of degree up to $15$ in $30$ variables with thousands of terms. So we had to combine several strategies that each reduced the time and/or memory complexity by one or two orders of magnitude. As much as possible, the code was written to be modular and reusable.

	\section{Context}

	Consider the polynomial ring $P=\QQ[X]$ with $k$ sets of $n$ variables. To fix conventions, we think of $X$ as a matrix of variables: \pauline{small $x$ for variables, capital $X$ for sets of variables? }
	$$(X_{i,j})_{0\leq i<k \atop 0\leq j<n}$$
	and we denote by $X_i$ the i-th set of variable.

	The polynomial ring $P$ is endowed with an action of the symmetric group $\Sn$ by diagonal permutation of the columns, and $\GLr$ by diagonal action on the rows. The invariant ring $P^{\Sn}$ is also known as the ring of multisymmetric functions. The co-invariant ring $P_{\Sn}$ is the quotient of $R$ by the ideal $\langle {P^{\Sn}}^+ \rangle$ \pauline{+ = without constant terms ?}. It inherits from $P$ a structure of graded $\GLr$-$\Sn$ bi-module.

	The grading can be refined into a multigrading by rows. The multigrading is preserved by the action of $\Sn$. The multigrading is not preserved by the action of $\GLr$, but instead carries exactly the information about the $\GLr$ character. It's thus equivalent to consider the $\GLr$-$\Sn$ bicharacter or the $\Sn$ multigraded character of $P_{\Sn}$.

	When $k=1$, the action of $GL_1$ is trivial; $P_{\Sn}$ is of dimension $n!$ and carries the usual graded representation of $\Sn$. Its character is given by a Hall-Littlewood polynomial.
	When $k=2$, $P_{\Sn}$ is of dimension $(n+1)^{n-1}$ ($=16807$ for $n=6$); as a consequence of Haiman's proof of the $n!$-conjecture, the bigraded character is a Macdonald polynomial.

	Throughout this notebook, we implicitly use the Frobenius map, and identify $\Sn$ and $\GLr$ characters with their associated symmetric functions. \\

	We denote the partial derivative the following way $\partial^k_x := \frac{\partial^k}{\partial x^k}$ for any variable $x$ in $X$.

	When $X$ contains only on set of variable ($r=1$), we define the \emph{harmonic polynomials} as follow.

	\begin{definition}
		Let $f \in \QQ[X]$, $f$ is \emph{harmonic} if and only if
		$$\partial^k_{x_1} f + \partial^k_{x_2} f + \dots + \partial^k_{x_n} f = 0 $$
		for all $k \geq 0$, $k \in \NN$.
	\end{definition}

	The space of harmonic polynomials in $n$ variables, denoted $\harmonics_{n}$ is the joint kernel of the differential operators $D_k := \partial_{x_1}^k + \cdots + \partial_{x_n}^k$. It is also a realization of the coinvariants of the symmetric group: $\QQ[X]\, /\, \langle Sym(X)^+ \rangle$ and a graded regular representation of $\Sn$.

	As an example, $\harmonics_2 = \langle 1, x_1-x_2 \rangle_\QQ$.

	Note that $\Delta:=$ $\prod_{i<j}(x_i-x_j)$ is harmonic and that if $f$ is harmonic then $\partial_i f$ is also harmonic. \\

	Let now go back to $X$ containing an arbitrary number $r$ of sets of variables.
	We define the \emph{differential operators} $(D_\alpha)_{\alpha\in \NN^r}$:
	$$D_{(\alpha_1, \dots, \alpha_r)} := \partial_{x_{11}}^{\alpha_1}\partial_{x_{r1}}^{\alpha_r} + \dots + \partial_{x_{1n}}^{\alpha_1}\partial_{x_{rn}}^{\alpha_r}$$

	For example, $r=3$ and we denote $x, y$ and $z$ the three sets of variables of $X$, then $D_{(3,0,4)} := \partial_{x_1}^3\partial_{z_1}^4 + \dots + \partial_{x_n}^3\partial_{z_n}^4$.

	The space of the diagonal harmonic polynomials $\harmonics_{n,r}$ is the joint kernel of the $(D_\alpha)_\alpha$. And $\harmonics_{n,r}$ contains $\harmonics_{n,1} = \harmonics_{n}$.

	%define polarization operators

@@

	We first need to compute in a quotient by an ideal.
	A standard approach would be to compute a Gröbner basis of this ideal.
	We haven't actually tried it, but given the number of variables and dimension of the quotient, we expect this calculation to be intractable.

@@

	Once a basis of $\harmonics_{k,n}$ is constructed, there still remains to recover its bicharacter. For $\GLr$ that's given by the multigrading. For $\Sn$ some additional work has to be done (e.g. computing the traces of the action of permutations of all cycle types).

@@

	Advantages:
	- Writing $N=:\sum N_D$, The complexity becomes $(\sum N_D^2) M$ instead of $(\sum N_D)^2M_D$ (in fact better.
	- The multidegrees, and therefore $\GLr$ character, can be directly read off the result.

@@

	The polarization operators commute with the action of $\Sn$. Therefore,
	we can compute independently each $\Sn$ isotypic component of $\harmonics_{k,n}$ by starting from a basis of harmonic polynomials that itself is adapted to its decomposition into isotypic components:
	$$\harmonics_{n}=:\bigoplus_\lambda \harmonics^\lambda\,.$$

@@

	Remark: For a fixed tableau $P$, denote by $\harmonics_k^P$ the space spanned by the higher Specht polynomials $(G_{P,Q})_Q$ under polarization. It's exactly the image of $\harmonics$ under the Young idempotent associated to $P$. For a given shape $\lambda$, the $\harmonics_k^P$ are all conjugate under $\Sn$, and their direct sum is the isotypic component $\harmonics_k^\lambda$.

@@

	\section{Applications}
